## Exercises 
1\. Compute conditional probabilities for being Male for the `heights` dataset.  Round the heights to the closest inch. Plot the estimated conditional probability $P(x) = \mbox{Pr}(\mbox{Male} | \mbox{height}=x)$ for each $x$.  
2\. In the plot we just made, we see high variability for low values of height. This is because we have few data points in these strata. This time use the `quantile` function for quantiles $0.1,0.2,\dots,0.9$ and the `cut` function to assure each group has the same number of points. Hint: for any numeric vector `x`, you can create groups based on quantiles like this:  
```{r, eval=FALSE} 
cut(x, quantile(x, seq(0, 1, 0.1)), include.lowest = TRUE) 
``` 
3\. Generate data from a bivariate normal distribution using the __MASS__ package like this: 
```{r, eval=FALSE} 
Sigma <- 9*matrix(c(1,0.5,0.5,1), 2, 2) 
dat <- MASS::mvrnorm(n = 10000, c(69, 69), Sigma) |> 
  data.frame() |> setNames(c("x", "y")) 
``` 
You can make a quick plot of the data using `plot(dat)`. Use an approach similar to the previous exercise to estimate the conditional expectations and make a plot. 

 
