## Exercises 
We are going to apply LDA and QDA to the `tissue_gene_expression` dataset. We will start with simple examples based on this dataset and then develop a realistic example. 
1\. Create a dataset with just the classes "cerebellum" and "hippocampus" (two parts of the brain) and a predictor matrix with 10 randomly selected columns. 
```{r, eval=FALSE} 
set.seed(1993) 
data("tissue_gene_expression") 
tissues <- c("cerebellum", "hippocampus") 
ind <- which(tissue_gene_expression$y %in% tissues) 
y <- droplevels(tissue_gene_expression$y[ind]) 
x <- tissue_gene_expression$x[ind, ] 
x <- x[, sample(ncol(x), 10)] 
``` 
Use the `train` function to estimate the accuracy of LDA. 
2\.  In this case, LDA fits two 10-dimensional normal distributions. Look at the fitted model by looking at the `finalModel` component of the result of train. Notice there is a component called `means` that includes the estimate `means` of both distributions. Plot the mean vectors against each other and determine which predictors (genes) appear to be driving the algorithm.  
3\. Repeat exercises 1 with QDA. Does it have a higher accuracy than LDA? 
4\. Are the same predictors (genes) driving the algorithm? Make a plot as in exercise 2. 
5\. One thing we see in the previous plot is that the value of predictors correlate in both groups: some predictors are low in both groups while others are high in both groups. The mean value of each predictor, `colMeans(x)`, is not informative or useful for prediction, and often for interpretation purposes it is useful to center or scale each column. This can be achieved with the `preProcessing` argument in `train`. Re-run LDA with `preProcessing = "scale"`. Note that accuracy does not change but see how it is easier to identify the predictors that differ more between groups in the plot made in exercise 4. 
6\. In the previous exercises we saw that both approaches worked well. Plot the predictor values for the two genes with the largest differences between the two groups in a scatterplot to see how they appear to follow a bivariate distribution as assumed by the LDA and QDA approaches. Color the points by the outcome. 
7\. Now we are going to increase the complexity of the challenge slightly: we will consider all the tissue types. 
```{r, eval=FALSE} 
set.seed(1993) 
data("tissue_gene_expression") 
y <- tissue_gene_expression$y 
x <- tissue_gene_expression$x 
x <- x[, sample(ncol(x), 10)] 
``` 
What accuracy do you get with LDA? 
8\. We see that the results are slightly worse. Use the `confusionMatrix` function to learn what type of errors we are making. 
9\. Plot an image of the centers of the seven 10-dimensional normal distributions. 

 
