## Exercises 
1\. In the wrangling part of this book, we used the code below to obtain mortality counts for Puerto Rico for 2015-2018. 
```{r, eval=FALSE} 
library(tidyverse) 
library(lubridate) 
library(purrr) 
library(pdftools) 
library(dslabs) 
fn <- system.file("extdata", "RD-Mortality-Report_2015-18-180531.pdf", 
                  package="dslabs") 
dat <- map_df(str_split(pdf_text(fn), "\n"), function(s){ 
  s <- str_trim(s) 
  header_index <- str_which(s, "2015")[1] 
  tmp <- str_split(s[header_index], "\\s+", simplify = TRUE) 
  month <- tmp[1] 
  header <- tmp[-1] 
  tail_index  <- str_which(s, "Total") 
  n <- str_count(s, "\\d+") 
  out <- c(1:header_index, which(n == 1),  
           which(n >= 28), tail_index:length(s)) 
  s[-out] |>  str_remove_all("[^\\d\\s]") |> str_trim() |> 
    str_split_fixed("\\s+", n = 6) |> .[,1:5] |> as_tibble() |>  
    setNames(c("day", header)) |> 
    mutate(month = month, day = as.numeric(day)) |> 
    pivot_longer(-c(day, month), names_to = "year", values_to = "deaths") |> 
    mutate(deaths = as.numeric(deaths)) 
}) |> 
  mutate(month = recode(month,  
                        "JAN" = 1, "FEB" = 2, "MAR" = 3,  
                        "APR" = 4, "MAY" = 5, "JUN" = 6,  
                        "JUL" = 7, "AGO" = 8, "SEP" = 9,  
                        "OCT" = 10, "NOV" = 11, "DEC" = 12)) |> 
  mutate(date = make_date(year, month, day)) |> 
  filter(date <= "2018-05-01") 
``` 
Use the `loess` function to obtain a smooth estimate of the expected number of deaths as a function of date. Plot this resulting smooth function. Make the span about two months long. 
2\. Plot the smooth estimates against day of the year, all on the same plot but with different colors. 
3\. Suppose we want to predict 2s and 7s in our `mnist_27` dataset with just the second covariate. Can we do this? On first inspection it appears the data does not have much predictive power. In fact, if we fit a regular logistic regression, the coefficient for `x_2` is not significant! 
```{r, eval = FALSE} 
library(broom) 
library(dslabs) 
data("mnist_27") 
mnist_27$train |>  
  glm(y ~ x_2, family = "binomial", data = _) |>  
  tidy() 
``` 
Plotting a scatterplot here is not useful since `y` is binary: 
```{r, eval = FALSE} 
qplot(x_2, y, data = mnist_27$train) 
``` 
Fit a loess line to the data above and plot the results. Notice that there is predictive power, except the conditional probability is not linear. 

 
