# ggplot2 {#ggplot2}

```{r, echo=FALSE}
img_path <- "R/img"
```

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```

## Motivation

"The greatest value of a picture is when it forces us to notice what we never expected to see."

John W. Tukey

## The components of a graph

* Data: The US murders data table is being summarized. We refer to this as the data component.

* Geometry: The plot above is a scatterplot. This is referred to as the geometry component. Other possible geometries are barplot, histogram, smooth densities, qqplot, and boxplot. We will learn more about these in the Data Visualization part of the book.

* Aesthetic mapping: The plot uses several visual cues to represent the information provided by the dataset. 

  - The two most important cues in this plot are the point positions on the x-axis and y-axis, which represent population size and the total number of murders, respectively. 
  - Each point represents a different observation, and we map data about these observations to visual cues like x- and y-scale. 
  - Color is another visual cue that we map to region. We refer to this as the aesthetic mapping component. 
  - The points are labeled with the state abbreviations.
  - How we define the mapping depends on what geometry we are using. 

We also have:


 
* The range of the x-axis and y-axis appears to be defined by the range of the data. 

* Both axes are in log-scales.

* There are labels, a title, a legend, and we use the style of The Economist magazine.


```{r}
library(dslabs)
data(murders)
```

## `ggplot` objects 

```{r ggplot-example-1, eval=FALSE}
ggplot(data = murders)
```


```{r ggplot-example-2}
murders |> ggplot()
```


```{r}
p <- ggplot(data = murders)
class(p)
```


```{r, eval=FALSE}
print(p)
p
```

## Geometries


>> DATA |> `ggplot()` + LAYER 1 + LAYER 2 + ... + LAYER N


Cheat sheet: https://github.com/rstudio/cheatsheets

<!--(Source: [RStudio](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf))-->


### geom_point help file:

```
> Aesthetics
> 
> geom_point understands the following aesthetics (required aesthetics are in bold):
>
> x
>
> y
> 
> alpha
>
> colour
```


## Aesthetic mappings

```{r, eval = FALSE}
murders |> ggplot() + 
  geom_point(aes(x = population/10^6, y = total))
```
 

Using already defined `p`

```{r ggplot-example-3}
p + geom_point(aes(population/10^6, total))
```


## Layers



```{r  ggplot-example-4}
p + geom_point(aes(population/10^6, total)) +
  geom_text(aes(population/10^6, total, label = abb))
```


```{r, eval=FALSE}
p_test <- p + geom_text(aes(population/10^6, total, label = abb))
```

Gives an error:

```{r, eval=FALSE}
p_test <- p + geom_text(aes(population/10^6, total), label = abb) 
```


### Tinkering with arguments
 

```{r ggplot-example-5}
p + geom_point(aes(population/10^6, total), size = 3) +
  geom_text(aes(population/10^6, total, label = abb))
```


```{r ggplot-example-6}
p + geom_point(aes(population/10^6, total), size = 3) +
  geom_text(aes(population/10^6, total, label = abb), nudge_x = 1.5)
```


## Global versus local aesthetic mappings


```{r}
args(ggplot)
```


```{r}
p <- murders |> ggplot(aes(population/10^6, total, label = abb))
```


```{r ggplot-example-7, eval=FALSE}
p + geom_point(size = 3) + 
  geom_text(nudge_x = 1.5)
```


```{r ggplot-example-8}
p + geom_point(size = 3) +  
  geom_text(aes(x = 10, y = 800, label = "Hello there!"))
```



## Scales



```{r ggplot-example-9}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") 
```


```{r, eval=FALSE}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() 
```


## Labels and titles

```{r ggplot-example-10}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")
```

## Categories as colors


```{r}
p <-  murders |> ggplot(aes(population/10^6, total, label = abb)) +   
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")
```


```{r ggplot-example-11}
p + geom_point(size = 3, color ="blue")
```

```{r ggplot-example-12}
p + geom_point(aes(col=region), size = 3)
```

## Annotation, shapes, and adjustments

```{r}
r <- murders |> 
  summarize(rate = sum(total) /  sum(population) * 10^6) |> 
  pull(rate)
```

Add intercept

```{r ggplot-example-13}
p + geom_point(aes(col=region), size = 3) + 
  geom_abline(intercept = log10(r))
```

change color

```{r}
p <- p + geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
  geom_point(aes(col=region), size = 3)  
```


```{r}
p <- p + scale_color_discrete(name = "Region") 
```

## Add-on packages {#add-on-packages}

```{r, eval}
ds_theme_set()
```

```{r, eval = FALSE}
library(ggthemes)
p + theme_economist()
```

Others:

```{r, eval = FALSE}
library(ggthemes)
p + theme_fivethirtyeight()
```


The add-on package __ggrepel__ includes a geometry that adds labels while ensuring that they don't fall on top of each other. We simply change `geom_text` with `geom_text_repel`.

## Putting it all together



```{r final-ggplot-example}
library(ggthemes)
library(ggrepel)

r <- murders |> 
  summarize(rate = sum(total) /  sum(population) * 10^6) |>
  pull(rate)

murders |> ggplot(aes(population/10^6, total, label = abb)) +   
  geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
  geom_point(aes(col=region), size = 3) +
  geom_text_repel() + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") + 
  scale_color_discrete(name = "Region") +
  theme_economist()
```

```{r, echo=FALSE}
ds_theme_set()
```


## Quick plots with `qplot` {#qplot}

```{r}
data(murders)
x <- log10(murders$population)
y <- murders$total
```


```{r, eval=FALSE}
data.frame(x = x, y = y) |> 
  ggplot(aes(x, y)) +
  geom_point()
```

```{r qplot-example-0, eval = FALSE}
qplot(x, y)
```


## Grids of plots


```{r gridExtra-example, warning=FALSE, message=FALSE, fig.height=2.5, fig.width=5}
library(gridExtra)
p1 <- qplot(x)
p2 <- qplot(x,y)
grid.arrange(p1, p2, ncol = 2)
```


## Exercises 

Start by loading the __dplyr__ and __ggplot2__ library as well as the `murders` and `heights` data.

```{r}
library(dplyr)
library(ggplot2)
library(dslabs)
data(heights)
data(murders)
```

1\. With __ggplot2__ plots can be saved as objects. For example we can associate a dataset with a plot object like this

```{r, eval = FALSE}
p <- ggplot(data = murders)
```

Because `data` is the first argument we don't need to spell it out

```{r, eval = FALSE}
p <- ggplot(murders)
```

and we can also use the pipe:

```{r, eval = FALSE}
p <- murders |> ggplot()
```

What is class of the object `p`?

 
2\. Remember that to print an object you can use the command `print` or simply type the object. 
Print the object `p` defined in exercise one and describe what you see.

a. Nothing happens.
b. A blank slate plot.
c. A scatterplot.
d. A histogram.


3\. Using the pipe `|>`, create an object `p` but this time associated with the `heights` dataset instead of the `murders` dataset. 

   

4\. What is the class of the object `p` you have just created?

   
5\. Now we are going to add a layer and the corresponding aesthetic mappings. For the murders data we plotted total murders versus population sizes. Explore the `murders` data frame to remind yourself what are the names for these two variables and select the correct answer. __Hint__: Look at `?murders`.

a. `state` and `abb`.
b. `total_murders` and `population_size`.
c. `total` and `population`.
d. `murders` and `size`.

6\. To create the scatterplot we add a layer with `geom_point`. The aesthetic mappings require us to define the x-axis and y-axis variables, respectively. So the code looks like this:

```{r, eval=FALSE}
murders |> ggplot(aes(x = , y = )) +
  geom_point()
```

except we have to define the two variables `x` and `y`. Fill this out with the correct variable names.


7\. Note that if we don't use argument names, we can obtain the same plot by making sure we enter the variable names in the right order like this:

```{r, eval=FALSE}
murders |> ggplot(aes(population, total)) +
  geom_point()
```

Remake the plot but now with total in the x-axis and population in the y-axis.


8\. If instead of points we want to add text, we can use the `geom_text()` or `geom_label()` geometries. The following code 

```{r, eval=FALSE}
murders |> ggplot(aes(population, total)) + geom_label()
```

will give us the error message: `Error: geom_label requires the following missing aesthetics: label`

Why is this?

a. We need to map a character to each point through the label argument in aes.
b. We need to let `geom_label` know what character to use in the plot.
c. The `geom_label` geometry does not require  x-axis and y-axis values.
d. `geom_label` is not a ggplot2 command.


9\. Rewrite the code above to use abbreviation as the label through `aes`


10\. Change the color of the labels to blue. How will we do this?

a. Adding a column called `blue` to `murders`.
b. Because each label needs a different color we map the colors through `aes`.
c. Use the `color` argument in `ggplot`.
d. Because we want all colors to be blue, we do not need to map colors, just use the color argument in `geom_label`.


11\. Rewrite the code above to make the labels blue.


12\. Now suppose we want to use color to represent the different regions. In this case which of the following is most appropriate:

a. Adding a column called `color` to `murders` with the color we want to use.
b. Because each label needs a different color we map the colors through the color argument of `aes` .
c. Use the `color` argument in `ggplot`.
d. Because we want all colors to be blue, we do not need to map colors, just use the color argument in `geom_label`.


13\. Rewrite the code above to make the labels' color be determined by the state's region.

14\. Now we are going to change the x-axis to a log scale to account for the fact the distribution of population is skewed. Let's start by defining an object `p` holding the plot we have made up to now

```{r, eval=FALSE}
p <- murders |> 
  ggplot(aes(population, total, label = abb, color = region)) +
  geom_label() 
```

To change the y-axis to a log scale we learned about the `scale_x_log10()` function. Add this layer to the object `p` to change the scale and render the plot.

15\. Repeat the previous exercise but now change both axes to be in the log scale.

16\. Now edit the code above to add the title "Gun murder data" to the plot. Hint: use the `ggtitle` function.
