---
title: "Dates And Times"
author: "Rafael A. Irizarry"
date: "`r lubridate::today()`"
output:
  ioslides_presentation:
    fig_caption: no
    fig_height: 5
    fig_width: 7
    out_width: "70%"
  beamer_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(dslabs)
library(gridExtra)
library(ggthemes)
ds_theme_set()
options(digits = 3)
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = "center",
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)

img_path <- "img"
```


## The date data type

-   We have described three main types of vectors: numeric, character, and logical.

-   In data science projects, we very often encounter variables that are dates.

-   Although we can represent a date with a string, for example `November 2, 2017`, once we pick a reference day, referred to as the *epoch*, they can be converted to numbers by calculating the number of days since the epoch.

-   Computer languages usually use January 1, 1970, as the epoch.

-   So, for example, January 2, 2017 is day 1, December 31, 1969 is day -1, and November 2, 2017, is day 17,204.

## The date data type

-   Now how should we represent dates and times when analyzing data in R? We could just use days since the epoch, but then it is almost impossible to interpret.

-   If I tell you it's November 2, 2017, you know what this means immediately.

-   If I tell you it's day 17,204, you will be quite confused.

-   Similar problems arise with times and even more complications can appear due to time zones.

-   For this reason, R defines a data type just for dates and times.


## The date data type

-   We see an example in the polls data:

```{r,  message=FALSE, warning=FALSE}
data("polls_us_election_2016") 
polls_us_election_2016$startdate |> head() 
```

-   These look like strings, but they are not:

```{r}
class(polls_us_election_2016$startdate) 
```

-   Look at what happens when we convert them to numbers:

```{r}
as.numeric(polls_us_election_2016$startdate) |> head() 
```

## The date data type

-   It turns them into days since the epoch.

-   The `as.Date` function can convert a character into a date.

-   So to see that the epoch is day 0 we can type.

```{r}
as.Date("1970-01-01") |> as.numeric() 
```

## The date data type

-   Plotting functions, such as those in ggplot, are aware of the date format.

-   This means that, for example, a scatterplot can use the numeric representation to decide on the position of the point, but include the string in the labels:

```{r rawpolls-vs-time, eval=FALSE}
polls_us_election_2016 |> 
  filter(pollster == "Ipsos" & state =="U.S.") |> 
  ggplot(aes(startdate, rawpoll_trump)) + 
  geom_line() 
```

## The date data type

```{r rawpolls-vs-time-2, echo=FALSE}
polls_us_election_2016 |> filter(pollster == "Ipsos" & state =="U.S.") |> 
  ggplot(aes(startdate, rawpoll_trump)) + 
  geom_line() 
```

-   Note that the month names are displayed, a very convenient feature. You can change this with a `scale_x_date` layer.

## The lubridate package

-   The **tidyverse** includes functionality for dealing with dates through the **lubridate** package.

```{r, warning=FALSE, message=FALSE}
library(lubridate) 
```

-   We will take a random sample of dates to show some of the useful things one can do:

```{r}
set.seed(2002) 
dates <- sample(polls_us_election_2016$startdate, 10) |> sort() 
dates 
```

## The lubridate package

-   The functions `year`, `month` and `day` extract those values:

```{r}
tibble(date = dates,  
       month = month(dates), 
       day = day(dates), 
       year = year(dates)) 
```

## The lubridate package

-   We can also extract the month labels:

```{r}
month(dates, label = TRUE) 
```

-   Another useful set of functions are the *parsers* that convert strings into dates.

## The lubridate package

-   The function `ymd` assumes the dates are in the format YYYY-MM-DD and tries to parse as well as possible.

```{r}
x <- c(20090101, "2009-01-02", "2009 01 03", "2009-1-4", 
       "2009-1, 5", "Created on 2009 1 6", "200901 !!! 07") 
ymd(x) 
```

-   A further complication comes from the fact that dates often come in different formats in which the order of year, month, and day are different.

## The lubridate package

-   The preferred format is to show year (with all four digits), month (two digits), and then day, or what is called the ISO 8601.

-   Specifically we use YYYY-MM-DD so that if we order the string, it will be ordered by date.

-   You can see the function `ymd` returns them in this format.

-   But, what if you encounter dates such as "09/01/02"? This could be September 1, 2002 or January 2, 2009 or January 9, 2002.

-   In these cases, examining the entire vector of dates will help you determine what format it is by process of elimination.

## The lubridate package

-   Once you know, you can use the many parses provided by **lubridate**.

-   For example, if the string is:

```{r}
x <- "09/01/02" 
```

## The lubridate package

-   The `ymd` function assumes the first entry is the year, the second is the month, and the third is the day, so it converts it to:

```{r}
ymd(x) 
```

-   The `mdy` function assumes the first entry is the month, then the day, then the year:

```{r}
mdy(x) 
```

## The lubridate package

-   The *lubridate* package provides a function for every possibility:

```{r}
ydm(x) 
myd(x) 
dmy(x) 
dym(x) 
```

## The lubridate package

-   The **lubridate** package is also useful for dealing with times.

-   In R base, you can get the current time typing `Sys.time()`.

-   The **lubridate** package provides a slightly more advanced function, `now`, that permits you to define the time zone:

```{r}
now() 
now("GMT") 
```

-   You can see all the available time zones with `OlsonNames()` function.

## The lubridate package

-   We can also extract hours, minutes, and seconds:

```{r}
now() |> hour() 
now() |> minute() 
now() |> second() 
```

## The lubridate package

-   The package also includes a function to parse strings into times as well as parsers for time objects that include dates:

```{r}
x <- c("12:34:56") 
hms(x) 
x <- "Nov/2/2012 12:34:56" 
mdy_hms(x) 
```

## The lubridate package

-   This package has many other useful functions.

-   We describe two of these here that we find particularly useful.

## The lubridate package

-   The `make_date` function can be used to quickly create a date object.

-   It takes three arguments: year, month, day, hour, minute, seconds, and time zone defaulting to the epoch values on UTC time.

-   So create an date object representing, for example, July 6, 2019 we write:

```{r}
make_date(2019, 7, 6) 
```

## The lubridate package

-   To make a vector of January 1 for the 80s we write:

```{r}
make_date(1980:1989) 
```

## The lubridate package

-   Another very useful function is the `round_date`.

-   It can be used to *round* dates to nearest year, quarter, month, week, day, hour, minutes, or seconds.

## The lubridate package

-   So if we want to group all the polls by week of the year we can do the following:

```{r poll-margin-versus-week-2, eval=FALSE}
polls_us_election_2016 |>  
  mutate(week = round_date(startdate, "week")) |> 
  group_by(week) |> 
  summarize(margin = mean(rawpoll_clinton - rawpoll_trump)) |> 
  ggplot(aes(week, margin)) + 
  geom_point() 
```

## The lubridate package

```{r poll-margin-versus-week, echo=FALSE}
polls_us_election_2016 |>  
  mutate(week = round_date(startdate, "week")) |> 
  group_by(week) |> 
  summarize(margin = mean(rawpoll_clinton - rawpoll_trump)) |> 
  ggplot(aes(week, margin)) + 
  geom_point() 
```
