## Exercises 
Project Gutenberg is a digital archive of public domain books. The R package __gutenbergr__ facilitates the importation of these texts into R. 
You can install and load by typing: 
```{r, eval=FALSE} 
install.packages("gutenbergr") 
library(gutenbergr) 
``` 
You can see the books that are available like this: 
```{r, eval=FALSE} 
gutenberg_metadata 
``` 
1\. Use `str_detect` to find the ID of the novel _Pride and Prejudice_.  
2\. We notice that there are several versions. The `gutenberg_works()` function filters this table to remove replicates and include only English language works. Read the help file and use this function to find the ID for _Pride and Prejudice_. 
3\. Use the `gutenberg_download` function to download the text for Pride and Prejudice. Save it to an object called `book`. 
4\. Use the __tidytext__ package to create a tidy table with all the words in the text. Save the table in an object called `words` 
5\. We will later make a plot of sentiment versus location in the book. For this, it will be useful to add a column with the word number to the table.  
6\. Remove the stop words and numbers from the `words` object. Hint: use the `anti_join`.  
7\. Now use the `AFINN` lexicon to assign a sentiment value to each word. 
8\. Make a plot of sentiment score versus location in the book and add a smoother. 
9\. Assume there are 300 words per page. Convert the locations to pages and then compute the average sentiment in each page. Plot that average score by page. Add a smoother that appears to go through data. 

 
