# Importing data {#importing-data}

## Exercises 


1\. Use the `read_csv` function to read each of the files that the following code saves in the `files` object: 

```{r, eval=FALSE}
path <- system.file("extdata", package = "dslabs")
files <- list.files(path)
files
```

  
2\. Note that the last one, the `olive` file, gives us a warning. This is because the first line of the file is missing the header for the first column. 

Read the help file for `read_csv` to figure out how to read in the file without reading this header. If you skip the header, you should not get this warning. Save the result to an object called `dat`.

  
3\. A problem with the previous approach is that we don't know what the columns represent. Type:

```{r, eval=FALSE}
names(dat)
```

to see that the names are not informative.

Use the `readLines` function to read in just the first line (we later learn how to extract values from the output). 



## Organizing data with spreadsheets

## Exercises 

1\. Pick a measurement you can take on a regular basis. For example, your daily weight or how long it takes you to run 5 miles. Keep a spreadsheet that includes the date, the hour, the measurement, and any other informative variable you think is worth keeping. Do this for 2 weeks. Then make a plot.


```{r, message=FALSE, echo=FALSE}
tmp <- file.remove("murders.csv")
```

